Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-02 analysis/paper/paper.Rmd
updates for introduction

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economi]. However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward. Also, specialization is not necessarily accompanied with social differentiation that is observed in lowland northern Philippines with low degree of ceramic specialization during the Spanish colonial period [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2018].

- standardization as indicator, metric and technological
- GM is more sensitive for subtle changes, and the outline
- innovative: CV test

craft specialization is commonly identified through statistical standardization of specific dimensions of a particular item – standardization as evidence of specialization.

The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Arnold2000, pp. 334; @Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, including metric, compositional, and technological approaches [@Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, linear measurements is widely used by calculation coefficient of variation (CV) that lower CV value indicates higher degree of standardization [@Hirshman2010; @Eerkens2001]. However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.

Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

- social changes -  culture contact framework 
- here I have the third cultural resistance, identity take a look at social boundary
- hypothesis, goals

This paper

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production. We observed pottery production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 600 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?

@@

## Geometric Morphometric approach: shape analysis for pottery specialization

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007]. Landmark-based morphometrics is widely used for archaeological objects where a morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

@@

- craft or pottery specialization
- CV is hard to test, we apply some statistics testing for it

>>>>

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic]. However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward. Also, specialization is not necessarily accompanied with social differentiation that is observed in lowland northern Philippines with low degree of ceramic specialization during the Spanish colonial period [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2018].

The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003].  However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic to identify changes in ceramic production in cross-cultural interaction contexts at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan. The aim of this paper is to investigate if there are increasing ceracmic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.

## Geometric Morphometric approach: shape analysis for ceramic specialization

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007]. Landmark-based morphometrics is widely used for archaeological objects where a morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

## social changes in cross-cultural interaction contexts
- here I have the third cultural resistance, identity take a look at social boundary
- hypothesis, goals

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production. We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 600 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?

@@

@@

- craft or pottery specialization
- CV is hard to test, we apply some statistics testing for it, - innovative: CV test
